Exploring the tree of numerical semigroups

### 2014-09-22 Papier/article.tex
Check of the paper + Wilf + Timings + Cilk/Timings + Abstract modif

\begin{abstract}
  In this paper we describe an algorithm visiting all the numerical semigroups
  up to a given genus using a new representation of numerical semigroups. The
  interest of this algorithm is that it fits particularly well the
  architecture of modern computer allowing very large optimizations.
\end{abstract}

@@

\section{Introduction}

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, close under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup[12,+\infty[
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of $\NN\setminus S$.
 For example the genus of $S_E$ is $6$,  the cardinality of $\{1,2,4,5,8,11\}$

@@

Here we describe a new algorithm for the exploration of the tree of numerical
semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroup that is well suited for exploration of the
tree~$\mathcal{T}$ and allow \Florent{large} code optimization essentially based on use of
vectorial instructions and parallelization. \Florent{The goal of the paper is twofold:
first to present our new encoding of the semigroup and the associated
algorithms, and second to present the optimization technique which allows for
those kinds of algorithms to get speedups by factors of hundreds and even
thousand. We claim that these technique are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores !}

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
\Jean{The decription of our new representation of numerical semigroup is done in the third section.}
\Delete{In Section~\ref{S:Tree} we describe a new representation of numerical semigrouos that is well suited to the construction of the tree.}
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimisation of the algorithm introduced in Section~\ref{S:Algo}.

@@

By definition a numerical semigroup is an infinite object. 
We need a finite description of such a semigroup. 
That is the role of generating sets.

@@

\begin{lem}[Lemma 2.3, Chap. I, \cite{BookNS}]
Let $S$ be a numerical semigroup. The set $\Irr(S)$ is the minimal generating set of $S$ relatively to the inclusion ordering.
\end{lem}

%\begin{proof}
%Assume for a contradiction, that there exists an integer  $x$ in $S$ than cannot be decomposed as a sum of irreducible elements. 
%We may assume that $x$ is  minimal with this property. 
%As  $x$ cannot be irreducible, there exist $y$ and $z$ in~$S\setminus\{0\}$ satisfying $x=y+z$. 
%Since we have $y<x$ and $z<x$, the integers $y$ and $z$ can be expressed as a sum of irreducible elements of $S$ and so $x=y+z$ is a sum of irreducible elements, in contradiction to hypothesis. 
%
%As irreducible elements of $S$ cannot be decomposed as the sum of two non-zeros integers in $S$, they must occur in each generating set of $S$.
%\end{proof}

%Let $S$ be a numericla semigroup. As $m(S)$ lies in $S$ the irreducible elements are include in
%\[
%\App(S)=\{x\in S \, |\, x-m(S)\not\in S\}
%\]
%The set $\App(S)$ is the so-called Ap\'ery set relativey to the element $m(S)$.
%
%\begin{lem}[Lemma 2.4, Chap. I, \cite{BookNS}]
%Let $S$ be a numerical semigroup. The Apéry set $\App(S)$ is finite of cardinality $m(S)$.
%\end{lem}
%
%The following result is a direct consequence of the previous Lemma and the Inclusion of $\Irr(S)$ in $m(S)$.
%
%\begin{prop}
%\label{P:Res}
%A numerical semigroup has at most $m(S)$ irreducible elements.
%\end{prop}

The differents parameters we have defined on a numerical semigroup, statisfy the following relations.

\begin{prop}
\label{P:Res}
Let $S$ be a numerical semigroup. We have

@@

$iii)$ Let $x$ be an element of $S\cap[0,c(S)-1]$. 
The integer $y=c(S)-1-x$ lies in $[0,c(S)-1]$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1\in S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:[0,c(S)-1]&\to&[0,c(S)-1]\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that send $S\cap[0,c(S)-1]$ into $[0,c(S)-1]\setminus S$ which is  of cardinality $g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap[0,c(S)-1]$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
[0,c(S)[=S\cap[0,c(S)[\sqcup[0,c(S)[\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

@@

We now explain the construction of the tree of numerical semigroups.
Let~$S$ be a numerical semigroup. The set $T=S\cup\{c(S)-1\}$ is also a numerical  semigroup and its genus is $g(S)-1$. 
As  $[c(S)-1,+\infty[$ is included in $T$ we have $c(T)\leq c(S)-1$. 
Therefore every semigroup $S$ of genus $g$ can be obtained from a semigroup $T$ of genus $g-1$ by removing an element of~$T$ greater than or equal to $c(T)$.

@@

Proposition~\ref{P:Sx} implies that every semigroup $T$ of genus $g$ can be obtained from a semigroup $S$ by removing a generator $x$ of $S$ that is greater than $c(S)$.
Hence relations~$T=S^x=S\setminus\{x\}$ hold.

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroup~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{\leq g}$ the subtree of $\mathcal{T}$ resticted to all semigroup of genus $\leq g$.

 \begin{figure}[hf!]
 % \[
 % \pstree[,nodesep=3pt]{\sgnode{1}}{
 % 	\pstree[nodesep=3pt]{\sgnode{2,3} \tlput1}{
 % 		\pstree[treesep=4pt,nodesep=2pt]{\sgnode{3,4,5} \tlput2}{
 % 			\pstree[treesep=3pt,nodesep=3pt]{\sgnode{4,5,6,7} \tlput3}{
 % 				\pstree{\sgnode{5,6,7,8,9} \tlput4}{}
 % 				\pstree{\sgnode{\gr 4,6,7,9} \tlput5}{}
 % 				\pstree{\sgnode{\gr 4,\gr 5,7} \trput6}{}
 % 				\pstree{\sgnode{\gr 4,\gr 5,\gr 6} \trput7}{}
 % 			}
 % 			\pstree[treesep=3pt,nodesep=3pt]{\sgnode{\gr 3,5,7} \tlput4}{
 % 				\pstree{\sgnode{\gr 3,7,8} \tlput 5}{}
 % 				\pstree{\sgnode{\gr 3,\gr 5} \trput 7}{}
 % 			}
 % 			\pstree[treesep=10pt]{\sgnode{\gr 3,\gr 4} \trput5}{\Tn \Tn \Tn}
 % 		}
 % 		\pstree[nodesep=3pt]{~\hspace{-5em}\sgnode{\gr2,5} \trput3}{
 % 			\pstree[nodesep=3pt]{\sgnode{\gr2,7} \trput5}{
 % 				\pstree[nodesep=3pt]{\sgnode{\gr2,9} \trput7}{}
 % 			}
 % 		}
 % 	}
 % }
 % \vspace{-5em}
 % \]

\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,
  level/.style={sibling distance=1.55cm},
%  edge from parent/.style={draw}
  ]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first four layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to $\mathcal{T}_{\leq 4}$. A generator of a semigroup is in gray if it is not in $[c(S),+\infty[$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ is $S'$ is obtained from $S$ by removing~$x$, that is $S'=S^x$.}
\label{F:Tree}
\end{figure}

As the reader can check, the main difficulty to characterize the son of a semigroup is to determine its irreducible elements.
 In \cite{NumericalSgps}, the semigroup are given by their Ap\'ery set and then the difficulty is to describe the Ap\'ery set of $S_x$  from this of $S$. 
 This approach is elegant but not sufficiently \emph{basic} for our optimisations.

@@

The aim of this section is to describe a new representation of numerical semigroups.
which is well suited to an efficient exploration of the tree~$\mathcal{T}$ of numerical semigroups.

@@

Assume that $y$ is an element of $D_S(x)$.
By very definition of $D_S(x)$, the integer~$z=x-y$ also belongs to $S$. 
Then $x$ can be decomposed as $x=y+z$ with $y$ and $z$ in~$S$. 
Moreover the condition $2y\leq x$ implies $y\leq z$. 
In other words if we define~$D'_S(x)$ to be the set of all $(y,z)\in S\times S$ with $x=y+z$ and $y\leq z$ then~$D_S(x)$ is the image of $D'_S(x)$ under the projection on the first coordinate.
 Hence~$D_S(x)$ describes how~$x$ can be decomposed as sums of two elements of $S$.
 This is we decide to call $d_S(x)$ the $S$-decomposition number of $x$.

@@

\begin{prop}
\label{P:Dandd}
Let $S$ be a numerical semigroup and $x\in\NN\setminus\{0\}$.  We have:

@@

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct three sets as follow:
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$.
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$ by Proposition~\ref{P:Dandd} the result holds.
\end{proof}

\section{A new algorithm}
\label{S:Algo}
We can easily explore the tree of numerical semigroups up to a genus $G$ using
a depth first search algorithm (see Algorithm~\ref{A:ExpRec}). This approach does not seem to
have been used before.  In particular, M. Bras-Amor\'os and M. Delgado use
instead a breadth first search exploration.  The main advantage in our
approach is the small memory needs.  The cost to pay is that, if we want to
explore the tree deeper, we must restart from the root.
\begin{algorithm}
\caption{Recursive Depth first search exploration of the tree of numerical semigroups}\label{A:ExpRec}
\begin{algorithmic}[1]
\Procedure{ExploreRec}{\tt{S}, \tt{G}}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State \Call{ExploreRec}{$\tt{S}^\tt{x}, \tt{G}$}
			\EndIf
		\EndFor
	\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

Equivalently, we can use an iterative version which uses a stack (see Algorithm~\ref{A:Exp}).
\begin{algorithm}
\caption{Iterative Depth first search exploration of the tree of numerical semigroups}\label{A:Exp}
\small
\begin{algorithmic}[1]
\Procedure{Explore}{\tt{G}}
\State Stack \tt{stack} \Comment{the empty stack}
\State $\tt{S} \gets\left<1\right>$
\While{stack is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State $\tt{S.push}(\tt{S}^\tt{x})$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have $c(S)\leq 2g(S)$ and so the $S$-decomposition number of $c(S)$ occurs in $\delta_S$.
Since $c(S)$ is equal to $\max(\NN\setminus S)$, Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in[0,...,3G],\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in[0,..,3G],\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in[0,...,3G],\ d_S(i)>0\}.
\]
Since, by Proposition~\ref{P:Res} $i)$,  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than $3G$ the relations $ii)$ and $iii)$ of Proposition~\ref{P:Res} give
\[
\Irr(S)=\{i\in[0,...,3G],\ d_S(i)=1\}.
\]
\end{proof}

@@

For example the following Algorithm initializes a representation of the semigroup $\NN$ ready for an exploration up to genus $G$.

\begin{algorithm}
\caption{Returns the root of $\mathcal{T}$ for an exploration up to genus \tt{G}}\label{A:Root}
\begin{algorithmic}
\small
\Function{Root}{\tt{G}}
\State \tt{R.c} $\gets$ 1 \Comment{\tt{R} stands for $\NN$}
\State \tt{R.g} $\gets$ 0
\State \tt{R.m} $\gets$ 1 
\For{\tt{x} from $1$ to $3\,\tt{G}$}
	\State\tt{R.d[x]}$\gets1+\left\lfloor\frac{\tt{x}}2\right\rfloor$
\EndFor
\State{\textbf{return} \tt{R}}
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Returns the son $S_x$ of $S$ with $x\in\Irr(S)\cap[c(S),c(S)+m(S)[$.}\label{A:Son}
\begin{algorithmic}[1]
\small
\Function{Son}{\tt{S},\tt{x},\tt{G}}
\State \tt{T.c} $\gets$ $\tt{x}+1$ \Comment{\tt{T}  stands for $\tt{S}_x$}
\State \tt{T.g} $\gets$ $\tt{S.g}+1$
\If{$\tt{x}>\tt{S.m}$}
	\State \tt{T.M} $\gets$ \tt{S.m}
\Else
	\State \tt{T.M} $\gets$ $\tt{S.m}+1$
\EndIf
\State\tt{T.d} $\gets$ \tt{S.d}
\For{\tt{y} from \tt{x} to $3\,\tt{G}$}
	\If{$\tt{S.d}[\tt{y}-\tt{x}]>0$}
		\State $\tt{T.d}[\tt{y}]\gets\tt{T.d}[\tt{y}]-1$
	\EndIf
\EndFor
\State{\textbf{return} \tt{T}}
\EndFunction
\end{algorithmic}
\end{algorithm}

>>>>

\begin{abstract}
  In this paper we describe an algorithm visiting all the numerical semigroups
  up to a given genus using a new representation of numerical semigroups. The
  interest of this algorithm is that it fits particularly well the
  architecture of modern computer allowing very large optimizations : we obtain the number of numerical semigroups of genus $g\leq  67$ and we confirm the Wilf conjecture for $g\leq 60$.

\end{abstract}

@@

\section*{Introduction}

A \emph{numerical semigroup} $S$ is a subset of $\NN$ containing $0$, close under addition and of finite complement in $\NN$.  
For example the set 
\begin{equation}
\label{E:NSG}
S_E=\{0,3,6,7,9,10\}\cup\{x\in\NN, x\geq 12\}
\end{equation}
is a numerical semigroup.
The \emph{genus} of a numerical semigroup $S$, denoted by~$g(S)$, is the cardinality of $\NN\setminus S$.
 For example the genus of $S_E$ is $6$,  the cardinality of $\{1,2,4,5,8,11\}$

@@

Here we describe a new algorithm for the exploration of the tree of numerical
semigroup $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroup that is well suited for exploration of the
tree~$\mathcal{T}$ and allow large code optimization essentially based on use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of the semigroup and the associated
algorithms, and second to present the optimization technique which allows for
those kinds of algorithms to get speedups by factors of hundreds and even
thousand. We claim that these technique are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores !

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
The decription of our new representation of numerical semigroup is done in the third section.
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimisation of the algorithm introduced in Section~\ref{S:Algo}.
Il the last section we emphase the results obtained using our algorithm.

@@

By definition a numerical semigroup is infinite object and we need a finite description of such a object. 
That is the role of generating sets.

@@

\begin{lem}[Lemma 2.3, Chap. I, \cite{BookNS}]
For a numerical semigroup $S$, the set $\Irr(S)$ is the minimal generating set of $S$ relatively to the inclusion ordering.
\end{lem}

The different parameters we have defined on a numerical semigroup, satisfy the following relations.

\begin{prop}
\label{P:Res}
For every  numerical semigroup  $S$,  we have

@@

$iii)$ Let $x$ be an element of $S\cap\{0,...,c(S)-1\}$. 
The integer $y=c(S)-1-x$ lies in $\{0,...,c(S)-1\}$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1\in S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:\{0,...,c(S)-1\}&\to&\{0,...,c(S)-1\}\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that send $S\{0,...,c(S)-1\}$ into $\{0,...,c(S)-1\}\setminus S$ which is  of cardinality $g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap\{0,...,c(S)-1\}$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
\{0,...,c(S)-1\}=S\cap\{0,...,c(S)-1\}\sqcup\{0,...,c(S)-1\}\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

@@

We now explain the construction of the tree of numerical semigroups.
Let~$S$ be a numerical semigroup. The set $S'=S\cup\{c(S)-1\}$ is also a numerical  semigroup and its genus is $g(S)-1$. 
As  each integer greater than $c(S)-1$ is included in $S'$ we have $c(S')\leq c(S)-1$. 
Therefore every semigroup $S$ of genus $g$ can be obtained from a semigroup $S'$ of genus $g-1$ by removing an element of~$S'$ greater than or equal to $c(S')$.

@@

Proposition~\ref{P:Sx} implies that every semigroup $S$ of genus $g$ can be obtained from a semigroup $S$ by removing a generator $x$ of $S$ that is greater than $c(S)$.
Hence the relation~$S'=S^x$ holds.

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroup~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{g}$ the subtree of $\mathcal{T}$ resticted to all semigroup of genus $\leq g$.

\vspace{-3em}

 \begin{figure}[hf!]
\[
\begin{tikzpicture}[level distance=2cm, inner sep=2mm,level/.style={sibling distance=1.55cm}]
    \node {$\sgnode{1}$}
    child {node {$\sgnode{2,3}$}
      child [sibling distance=4.5cm] {node {$\sgnode{3,4,5}$}
        child {node {$\sgnode{4,5,6,7}$}
          child [sibling distance=2cm]{node {$\sgnode{5,6,7,8,9}$}  edge from parent node [left] {4}}
          child [sibling distance=1.9cm]{node {$\sgnode{\gr 4,6,7,9}$}  edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 4,\gr 5,7}$}  edge from parent node [right] {6}}
          child {node {$\sgnode{\gr 4,\gr 5,\gr 6}$}  edge from parent node [right] {7}}
          edge from parent node [left] {3}
        }
        child{{} edge from parent[draw=none]}
        child{{} edge from parent[draw=none]}
        child {node {$\sgnode{\gr 3,5,7}$}
          child {node {$\sgnode{\gr 3,7,8}$} edge from parent node [left] {5}}
          child {node {$\sgnode{\gr 3,\gr 5}$} edge from parent node [right] {7}}
          edge from parent node [left] {4}
        }
        child {node {$\sgnode{\gr 3,\gr 4}$}
          edge from parent node [right] {5}
        }
        edge from parent node [left] {2}
      }
      child [sibling distance=4.5cm] {node {$\sgnode{\gr2,5}$}
        child {node {$\sgnode{\gr2,7}$}
          child {node {$\sgnode{\gr2,9}$}
            edge from parent node [right] {7}
          }
          edge from parent node [right] {5}
        }
        edge from parent node [right] {3}
      }
      edge from parent node [left] {1}
    }
    ;
  \end{tikzpicture}
  \]
\caption{The first four layers of the tree $\mathcal{T}$ of numerical semigroups, corresponding to $\mathcal{T}_4$. A generator of a semigroup is in gray if is not greater that $c(S)$. An edge between a semigroup $S$ and its son $S'$ is labelled by  $x$ is $S'$ is obtained from $S$ by removing~$x$, that is $S'=S^x$.}
\label{F:Tree}
\end{figure}

As the reader can check, the main difficulty to characterize the son of a semigroup is to determine its irreducible elements.
 In \cite{NumericalSgps}, the semigroup are given by their Ap\'ery set and then the difficulty is to describe the Ap\'ery set of $S^x$  from this of $S$. 
 This approach is elegant but not sufficiently \emph{basic} for our optimisations.

@@

The aim of this section is to describe a new representation of numerical semigroups,
which is well suited to an efficient exploration of the tree~$\mathcal{T}$ of numerical semigroups.

@@

Assume that $y$ is an element of $D_S(x)$.
By very definition of $D_S(x)$, the integer~$z=x-y$ also belongs to $S$. 
Then $x$ can be decomposed as $x=y+z$ with $y$ and $z$ in~$S$. 
Moreover the condition $2y\leq x$ implies $y\leq z$. 
In other words if we define~$D'_S(x)$ to be the set of all $(y,z)\in S\times S$ with $x=y+z$ and $y\leq z$ then~$D_S(x)$ is the image of $D'_S(x)$ under the projection on the first coordinate.
 Hence~$D_S(x)$ describes how~$x$ can be decomposed as sums of two elements of $S$.
 This is why we decide to call $d_S(x)$ the $S$-decomposition number of $x$.

@@

\begin{prop}
\label{P:Dandd}
For a numerical semigroup $S$ and $x\in\NN\setminus\{0\}$, we have:

@@

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct two sets as follow:
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$, by Proposition~\ref{P:Dandd} the result holds.
\end{proof}

\section{A new algorithm}
\label{S:Algo}
We can easily explore the tree of numerical semigroups up to a genus~$G$ using
a depth first search algorithm (see Algorithm~\ref{A:ExpRec}). This approach does not seem to
have been used before.  In particular, M. Bras-Amor\'os and M.~Delgado use
instead a breadth first search exploration.  The main advantage in our
approach is the small memory needs.  The cost to pay is that, if we want to
explore the tree deeper, we must restart from the root.

\begin{algorithm}
\small
\caption{\small Recursive Depth first search exploration of the tree $\mathcal{T}_g$.}\label{A:ExpRec}
\begin{algorithmic}[1]
\Procedure{ExploreRec}{\tt{S}, \tt{G}}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State \Call{ExploreRec}{$\tt{S}^\tt{x}, \tt{G}$}
			\EndIf
		\EndFor
	\EndIf
\EndProcedure
\end{algorithmic}
\end{algorithm}

Equivalently, we can use an iterative version which uses a stack:
\begin{algorithm}
\caption{\small Iterative Depth first search exploration of the tree $\mathcal{T}_g$}\label{A:Exp}
\small
\begin{algorithmic}[1]
\Procedure{Explore}{\tt{G}}
\State Stack \tt{stack} \Comment{the empty stack}
\State $\tt{S} \gets\left<1\right>$
\While{stack is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State $\tt{S.push}(\tt{S}^\tt{x})$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have  the relation $c(S)\leq 2g(S)$ and so the $S$-decomposition number of $c(S)$ occurs in $\delta_S$.
Since $c(S)$ is equal to $\max(\NN\setminus S)$, Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in\{0,...,3G\},\ d_S(i)>0\}.
\]
Since, by Proposition~\ref{P:Res} $i)$,  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than $3G$ from the relations $ii)$ and $iii)$ of Proposition~\ref{P:Res}. So Proposition~\ref{P:Dandd} gives
\[
\Irr(S)=\{i \in\{0,...,3G\},\ d_S(i)=1\}.
\]
\end{proof}

@@

For example the following Algorithm initializes a representation of the semigroup $\NN$ ready for an exploration of the tree $\mathcal{T}_G$, \ie, the tree of numerical semigroup upto genus $G$.

\begin{algorithm}
\caption{\small Returns the root of $\mathcal{T}_G$}\label{A:Root}
\begin{algorithmic}
\small
\Function{Root}{\tt{G}}
\State \tt{R.c} $\gets$ 1 \Comment{\tt{R} stands for $\NN$}
\State \tt{R.g} $\gets$ 0
\State \tt{R.m} $\gets$ 1 
\For{\tt{x} from $1$ to $3\,\tt{G}$}
	\State\tt{R.d[x]}$\gets1+\left\lfloor\frac{\tt{x}}2\right\rfloor$
\EndFor
\State{\textbf{return} \tt{R}}
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\newpage
